From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study

### 2025-08-25 draft/draft_v3.tex
add draft_v3 (nature format)

Mental health disorders represent a growing global health concern,
especially in urbanized regions (Gruebner et al., 2017; World Health
Organization, 2022). In 2019, one in eight people worldwide were living
with a mental health condition, with the highest disability-adjusted
life years (DALYs) due to mental and addictive disorders present in
high-income countries such as those in Northern Europe, North America,
and Australia (Rehm \& Shield, 2019; World Health Organization, 2022).
Despite growing awareness, governments have struggled to design
effective responses. Mental health outcomes arise from complex,
multi-level dynamics between the social, economic, and spatial
structures of urban environments---making both diagnosis and
intervention design deeply challenging (Van Der Wal et al., 2021).

Recent work has shifted focus from individual-level vulnerability to
broader social and structural determinants. While high-income nations
show elevated overall DALYs, within-country disparities reveal that
income inequality is a significant predictor of mental health burden
(Rehm \& Shield, 2019). Moreover, precarious working conditions, housing
instability, and neighborhood disadvantage have all been linked to
increased risk of depression and anxiety (Fone et al., 2014; Pevalin et
al., 2017; Rönnblad et al., 2019; Rugulies et al., 2023). This growing
literature highlights that poor mental health is not merely a personal
issue but one rooted in sustained socioeconomic stress and structural
uncertainty.

Building on this, recent studies have introduced the concept of
precariousness as a multidimensional condition characterized by
instability and lack of control across several life domains, including
employment, finances, housing, social relations, and cultural belonging
(Elsenburg et al., 2025; McKee et al., 2017). This broader view reveals
how different forms of insecurity can co-occur and compound, forming an
ecosystem of risk. However, despite growing evidence of association
between precariousness and mental health, a fundamental question remains
unresolved: \emph{How do these factors causally influence one another?}
Most existing studies rely on cross-sectional correlations, leaving open
questions about directionality and feedback. Without stronger causal
insight, it remains difficult to identify priority targets for
intervention or to anticipate and quantify intended and unintended
effects.

This study aims to investigate the causal mechanisms linking
precariousness and depression and how those mechanisms shape system
responses to external intervention. Specifically, we ask: \emph{How does
the internal configuration of the precariousness--depression
relationship affect the outcomes of stress-reducing interventions?} To
answer this, we combine two approaches. First, we use cycle-capable
causal discovery algorithms to infer directional relationships among
domain-specific precariousness factors and depressive symptoms. By
examining both aggregated measures (depression sum score and a composite
precariousness index) and disaggregated variables (individual depressive
symptoms and specific life-domain indicators), we capture broad patterns
of relationship as well as fine-grained causal pathways. Second, we
translate these insights into a computational dynamical model that
simulates how depression and precariousness co-evolve under varying
levels of external stress. This model enables us to probe how system
properties---such as feedback strength and stochastic noise---govern
responsiveness to intervention. When the effects of depression symptoms
onto the social precarity subsystem are weaker than the reverse path,
there is higher sensitivity of depression to intervention, reflecting
both a faster reduction post financial intervention but also quicker
relapse once financial stresses are reestablished. By integrating causal
discovery with dynamic simulation, we demonstrate that intervention
strategies can be tailored to the underlying architecture of
vulnerability.

\section{Methods}\label{methods}

We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study
(Snijder et al., 2017; Stronks et al., 2013). It captures the diverse
population of the city of Amsterdam by including the six main ethnic
groups and provides comprehensive health and lifestyle data, including
depressive symptoms as measured by the PHQ-9 (Galenkamp et al., 2017).
To operationalize indicators of precariousness, we draw on the framework
outlined in previous research (Elsenburg et al., 2025) and select a set
of relevant variables. To ensure a robust representation of
precariousness in our causal discovery models, we conducted exploratory
analyses to identify consistent and meaningful data-driven factor
structures. These analyses led to the identification of five distinct
precariousness factors: three reflecting structural
domains---employment, housing, and social relationships---and two
capturing recent life stressors---financial and relational. Each factor
is composed of multiple observed variables, as outlined below.
Additional details on the HELIUS study and the factor extraction process
are provided in the \hyperref[sec-appendix]{Appendix}.

Following data cleaning and normalization procedures, the analytical
sample comprises 21,628 individuals. In addition to the five identified
precariousness factors, we computed an overall precariousness score by
summing these five factor scores, thereby capturing a composite measure
of accumulated precariousness across domains. Depression is represented
using the PHQ-9 (Kroenke et al., 2001), both as a total sum score and as
individual symptom scores. In the subsequent causal discovery analysis,
we explore the relationship between depression and precariousness using
both aggregated scores and their disaggregated components. See
Figure~\ref{fig-dist} for the overall distributions of the variables
used in the analysis.

\includegraphics[width=0.9\textwidth,height=\textheight]{draft_v2_files/figure-pdf/fig-dist-1.pdf}

To uncover directional relationships between precariousness and
depression, we use constraint-based causal discovery algorithms that
allow for feedback loops and latent confounding. Specifically, we apply
the \emph{Fast Causal Inference} (FCI) and \emph{Cyclic Causal
Inference} (CCI) algorithms, both of which can detect cycles and
non-acyclic structures (Mooij \& Claassen, 2020; Strobl, 2019). For
reference, we also include the \emph{PC} algorithm, one of the most
widely used methods for acyclic causal discovery (Spirtes et al., 2001).
While our full analysis includes results from three causal discovery
algorithms---FCI, CCI, and PC---we focus on FCI in the main text, as it
offers a strong balance between flexibility and theoretical soundness.
FCI accommodates latent confounding and potential feedback, making it
well-suited for the complexity of our data (Mooij \& Claassen, 2020).
The PC algorithm is included for reference but relies on more
restrictive assumptions, such as acyclicity and no unmeasured
confounding. CCI, while capable of modeling cycles explicitly, has its
own theoretical limitations; for example, its output may not fully
preserve the d-separation properties of the true causal graph (Strobl,
2019). Results from both PC and CCI are provided in the
\hyperref[sec-appendix]{Appendix}
(Section~\ref{sec-cci}, Section~\ref{sec-pc}) for completeness and
comparison. For readers interested in detailed explanations of the
algorithms, edge interpretations, and graph types (e.g., PAG, MAAG,
CPDAG), as well as algorithmic assumptions and limitations, please refer
to the \hyperref[sec-appendix]{Appendix}
(Section~\ref{sec-causalprimer}).

A key challenge in applying these algorithms to the HELIUS dataset is
that many variables are non-Gaussian and likely exhibit nonlinear
relationships. While kernel-based conditional independence tests (such
as KCIT) are well-suited for capturing such complex dependencies, they
are computationally intensive---scaling quadratically with sample size
due to the need to invert large kernel matrices---making them
impractical for large datasets like HELIUS (Rahimi \& Recht, 2007). To
address this, we use the Randomized Conditional Correlation Test (RCoT),
a nonparametric alternative that approximates kernel methods using
random Fourier features. This reduces computational complexity from
quadratic to linear in sample size, substantially lowering runtime while
preserving sensitivity to nonlinear relationships (Strobl et al., 2019;
Zhang et al., 2012). For further details, see Section~\ref{sec-rcot}.

@@

\caption{\label{fig-concept}High-level schematic of the dynamic model
structure used in simulation. Arrows represent influence coefficients
\(\alpha\), matching the terms in the system of stochastic differential
equations. \(D_i(t)\) and \(P_i(t)\) denote depressive symptoms and
social precariousness for individual \(i\), with \(S_i\) representing
static financial stress.}

@@

@@

\caption{\label{fig-intervention}System response to external
intervention across variations in internal structure.
\emph{\textbf{Panel A}: Mean levels of depression (\(D\)) and
precariousness (\(P\)) at time steps 100, 300, and 500, across varying
values of \(\alpha_{DP}\) and intervention strengths. The x-axis
represents the strength of external support, operationalized as the
proportional reduction of financial stress (\(S\)) toward its minimum
observed value. Each line corresponds to a different model defined by a
specific \(\alpha_{DP}\) value, color-coded from low (purple) to high
(yellow). Models with lower \(\alpha_{DP}\) exhibit faster early
reductions in \(D\), while higher \(\alpha_{DP}\) leads to more
immediate declines in \(P\). \textbf{Panel B}: Full time trajectories
for two models---one with no coupling (\(\alpha_{DP} = 0\)) and one with
strong coupling (\(\alpha_{DP} = 0.65\))---illustrating system response
and recovery following intervention onset (step 300) and withdrawal
(step 1500). Depression responds more gradually and persistently in the
high \(\alpha_{DP}\) model, whereas precariousness adjusts more
quickly.}}

@@

This study combined constraint-based causal discovery and dynamical
modeling to investigate how depression and socioeconomic precariousness
interact and how their internal structure modulates responsiveness to
intervention. Consistent with prior work linking financial strain to
psychological distress (Ezzy, 1993; Lund et al., 2010), our analysis
identified financial stress as a key upstream driver of depression in
both aggregated and disaggregated models. This supports longstanding
hypotheses that economic stressors can act as modifiable entry points
into broader cycles of psychosocial risk.

At the symptom level, our analysis aligns with established theories of
psychopathology but provides a more fine-grained map of potential causal
relationships, generating concrete, testable hypotheses. It also points
to possible limitations of prevailing models, as some observed
symptom--symptom links may be shaped by shared confounders rather than
direct causal influence. This suggests that, in some contexts, broader
mechanisms or clusters of symptoms may provide more effective units of
analysis and intervention than individual symptoms alone. Within this
framework, sleep disturbance emerged as a potentially initiating
symptom, with directed connections to both symptom-level and
precariousness variables---mirroring longitudinal findings that identify
sleep disruption as an early warning signal for depressive episodes and
social withdrawal (Alvaro et al., 2013; Zawadzki et al., 2013).
Depressed mood, another core symptom, occupied a central position in the
network. Not only was it influenced by upstream symptoms like anhedonia
and guilt, but it also showed directional influence on social
precariousness---suggesting that internal psychological states may
actively shape external vulnerability. The CCI algorithm further
identified a potential feedback loop between depressed mood and overall
precariousness, resonating with social drift theories in depression,
which posit that declining mental health can lead to social exclusion,
job instability, and housing insecurity, reinforcing depressive cycles
(Boschloo et al., 2015; Compton \& Shim, 2015).

In contrast, social precariousness itself appeared more often as an
effect than a cause, receiving arrows from symptoms like depressed mood
and guilt but rarely sending them. This asymmetry supports prior
observations that indicators of social vulnerability often reflect
downstream consequences of poor mental health, though the relationships
are likely bidirectional over time (Lund et al., 2010). Although our
findings are based on Amsterdam-specific data, they align with research
in other high-income urban settings, where structural inequalities such
as poverty, unemployment, and housing insecurity consistently predict
poor psychological outcomes (Fryers et al., 2003; Weich \& Lewis, 1998).
While local contexts may alter specific pathways, these parallels
suggest that similar mechanisms could operate elsewhere, though
replication in other populations is needed to assess robustness.

Taken together, these findings yield actionable insights for mental
health intervention. Symptoms like sleep disturbance, which demonstrate
early and broad influence across both symptom and precariousness
networks, may serve as effective targets for prevention---potentially
halting the progression into more entrenched depressive states or
worsening precariousness. Similarly, targeting symptoms such as
depressed mood, which appear embedded in potential feedback loops
linking psychological and social domains, may offer strategic leverage
for breaking mutually reinforcing cycles of mental decline and social
isolation. Importantly, interventions targeting such nodes may have
ripple effects across domains. For instance, evidence suggests that
treating core depressive symptoms can improve social functioning,
employment engagement, and housing stability---particularly when
combined with social or financial supports (Lund et al., 2018; McDaid et
al., 2019). These dual benefits underscore the value of integrated,
cross-domain strategies that simultaneously address internal symptoms
and external stressors.

Our results also highlight why single-point interventions may fail to
produce lasting change. The broader causal structure reveals multiple
interlocking feedback loops spanning both psychological symptoms and
structural risk factors, which may re-amplify vulnerability even after
temporary relief. Supporting this, recent simulation work by Park, Li,
et al. (2025) shows that feedback loops, even when modest in strength,
can substantially diminish the long-term impact of isolated
interventions unless multiple reinforcing links are simultaneously
disrupted. Our own simulations support this view: while all models
converged to similar long-term outcomes under sustained support,
differences in internal structure---specifically, the strength of the
link from precariousness to depression---produced systematic variation
in how quickly and through which pathways systems responded to
intervention. Stronger coupling slowed recovery and prolonged
depression, showing that internal architecture can be as important as
external input. This supports growing calls for systems-oriented mental
health perspectives (Borsboom, 2017). Rather than assuming uniform
responses, more durable outcomes may require (1) targeting early-acting
symptoms, (2) sustaining support where recovery is slow, and (3)
combining structural and psychological interventions. Tailoring
strategies to match the system's architecture can improve effectiveness,
particularly in populations with layered vulnerabilities. In this light,
understanding causal architecture is not merely theoretical---it is
foundational to designing more effective and equitable interventions.

While the FCI and CCI algorithms are designed to recover causal
structure from cross-sectional data---including feedback and latent
confounding---many edge directions in our results remain ambiguous,
particularly in links between individual symptoms and precariousness
factors. These inconsistencies likely reflect limited statistical signal
in high-dimensional settings, where large conditioning sets reduce power
and small effect sizes are harder to detect. Moreover, our findings
showed sensitivity to the choice of conditional independence (CI) test.
The Gaussian CI test, which assumes linearity and Gaussian noise, often
produced denser graphs, likely due to overestimating weak or noisy
linear associations. In contrast, RCoT, a nonparametric alternative,
avoids these assumptions but relies on a Gaussian radial basis function
(RBF) kernel to assess dependence. While flexible in capturing nonlinear
relationships, this kernel is optimized for smooth, continuous variables
and may underperform with discrete or ordinal data. Its similarity
estimates can be unreliable when applied to variables with
discontinuities, potentially leading to missed dependencies. This is
particularly relevant in the HELIUS dataset, where some variables are
ordinal or mixed-type (Howlett, 2001). To address this, future work
should consider adapting kernel-based CI tests to better handle mixed
data structures---for example, by employing hybrid kernels that
explicitly combine continuous and discrete similarity measures. This
could improve the reliability of nonparametric causal discovery in
settings with heterogeneous variable types.

Although the CCI algorithm is explicitly designed to detect cycles,
aside from the connection between precarity and depression in the
aggregate precarity setting, we found little clear evidence of feedback:
most candidate loops appeared only as bidirectional edges, which may
reflect latent confounding or statistical imprecision. This could point
to either an absence of strong feedback at the symptom and the
precariousness factor levels or limitations in the sensitivity of
full-graph discovery approaches applied to densely interconnected
variables. More focused strategies, such as local structure tests or
theory-guided subgraph modeling, may help resolve these ambiguities.
Moreover, our analyses relied solely on observational data from the
Amsterdam population. Some structural uncertainties, particularly those
involving latent confounding or bidirectional influence, may ultimately
require interventional or semi-interventional designs. Extensions such
as LLC (Hyttinen et al., 2012), NODAGS-Flow (Sethuraman et al., 2023),
or Bicycle (Rohbeck et al., 2024), which integrate observational and
interventional data, could be especially valuable where such data are
available.

Finally, the structure of our dynamical model introduces important
limitations. The model focused on a restricted set of variables
(depression, social precariousness, and financial stress). This design
was intentionally simplified to isolate core mechanisms of interest, but
it necessarily omits other relevant domains of precariousness,
additional stressors, and potential unintended consequences of
intervention. In addition, while the linear formulation offered
analytical tractability and transparent calibration, it likely
oversimplifies the true dynamics of depression and socioeconomic
precariousness. A growing body of evidence suggests that psychosocial
systems are fundamentally nonlinear. Empirical and theoretical work
points to phenomena such as threshold effects, where improvements only
occur after a tipping point is crossed (Leemput et al., 2014; Park,
Waldorp, et al., 2025); saturation, where the impact of support
diminishes as basic needs are met (Cramer et al., 2010); and path
dependence, where past adversity shapes sensitivity to future stressors
(Wichers et al., 2016). These behaviors are fundamentally incompatible
with linear dynamics, yet they may be essential for understanding why
some individuals remain trapped in high-risk states while others recover
more easily. Future modeling efforts should therefore incorporate
mechanisms such as saturating input-response functions (e.g., sigmoidal
form) and state-dependent feedback (where coupling strength varies by
system state). Incorporating such features would allow models to capture
not just average effects but also the critical conditions under which
meaningful change becomes possible---clarifying whether some subgroups
require forceful, threshold-crossing interventions, while others may
benefit from sustained, moderate support.

As public health research has long shown, social and psychological
systems do not always respond passively to external inputs; internal
constraints can limit the impact of even well-intentioned policies
(Savigny \& Adam, 2009; Wagenaar \& Burris, 2013). By using causal
discovery and computational modeling, researchers and policymakers can
better identify these structural constraints and tailor interventions
accordingly. More broadly, this study offers a generalizable framework
for investigating complex mental health dynamics. Causal discovery
supports principled hypothesis generation from observational data, while
dynamical modeling enables exploration of how hypothesized structures
behave under external change. Together, these tools move us beyond
static descriptions toward mechanistic insight with direct relevance for
real-world mental health policy and intervention design. Future research
should build on this foundation by developing richer, more nuanced
dynamical models grounded in longitudinal and interventional data.
Incorporating time, multiple domains of precarity, and adaptive feedback
processes would help assess whether the mechanisms proposed manifest in
actual mental health trajectories. Ultimately, we hope this study
encourages further integration of causal structure learning and systems
modeling in efforts to understand and improve mental health outcomes
under conditions of socioeconomic precariousness.

>>>>

Mental health disorders represent a growing global health concern,
especially in urbanized regions\textsuperscript{1,2}. In 2019, one in
eight people worldwide were living with a mental health condition, with
the highest disability-adjusted life years (DALYs) due to mental and
addictive disorders present in high-income countries such as those in
Northern Europe, North America, and Australia\textsuperscript{1,3}.
Despite growing awareness, governments have struggled to design
effective responses. Mental health outcomes arise from complex,
multi-level dynamics between the social, economic, and spatial
structures of urban environments---making both diagnosis and
intervention design deeply challenging\textsuperscript{4}.

Recent work has shifted focus from individual-level vulnerability to
broader social and structural determinants. While high-income nations
show elevated overall DALYs, within-country disparities reveal that
income inequality is a significant predictor of mental health
burden\textsuperscript{3}. Moreover, precarious working conditions,
housing instability, and neighborhood disadvantage have all been linked
to increased risk of depression and anxiety\textsuperscript{5--8}. This
growing literature highlights that poor mental health is not merely a
personal issue but one rooted in sustained socioeconomic stress and
structural uncertainty.

Building on this, recent studies have introduced the concept of
precariousness as a multidimensional condition characterized by
instability and lack of control across several life domains, including
employment, finances, housing, social relations, and cultural
belonging\textsuperscript{9,10}. This broader view reveals how different
forms of insecurity can co-occur and compound, forming an ecosystem of
risk. However, despite growing evidence of association between
precariousness and mental health, a fundamental question remains
unresolved: \emph{How do these factors causally influence one another?}
Most existing studies rely on cross-sectional correlations, leaving open
questions about directionality and feedback. Without stronger causal
insight, it remains difficult to identify priority targets for
intervention or to anticipate and quantify intended and unintended
effects.

This study aims to investigate the causal mechanisms linking
precariousness and depression and how those mechanisms shape system
responses to external intervention. Specifically, we ask: \emph{How does
the internal configuration of the precariousness--depression
relationship affect the outcomes of stress-reducing interventions?} To
address this question, we analyze cross-sectional data from more than
21,000 participants in the HELIUS (HEalthy LIfe in an Urban Setting)
study\textsuperscript{11,12} using a two-step approach. First, we apply
cycle-capable causal discovery algorithms to infer directional
relationships among financial stress, domain-specific precariousness,
and depressive symptoms. By examining both aggregated measures (a
depression sum score and a composite precariousness index) and
disaggregated indicators (individual symptoms and life-domain factors),
we capture both broad patterns and fine-grained pathways. Second, we
translate these insights into a computational dynamical model that
simulates how depression and precariousness co-evolve under varying
levels of external stress. This framework enables us to examine how
system properties---such as feedback strength and stochastic
noise---shape responsiveness to intervention.

Figure~\ref{fig-concept} presents this high-level model, derived from
the causal discovery analysis and serving as the foundation for our
simulation experiments. It highlights financial stress (\(S_i\)) as an
upstream driver influencing both depressive symptoms (\(D_i\)) and
social precariousness (\(P_i\)). Depression contributes to broader
social vulnerability by feeding into precariousness, while
precariousness feeds back into depression, creating reinforcing cycles.
This framework illustrates that intervention effects depend not only on
external inputs but also on the internal feedback architecture of the
system. Full methodological details, including data processing,
algorithmic specifications, and model calibration procedures, are
provided in the \hyperref[sec-methods]{Methods} section.

@@

\caption{\label{fig-concept}The model abstracts the main dependencies
revealed by causal discovery into a compact system of dynamic
interactions. Financial stress (\(S_i\)) is treated as an upstream
driver that influences both depressive symptoms (\(D_i(t)\)) and social
precariousness (\(P_i(t)\)). Depression and precariousness also
influence each other through feedback loops, creating the potential for
reinforcing cycles of vulnerability. Arrows represent influence
coefficients (\(\alpha\)), which correspond to terms in the system of
stochastic differential equations used for simulation. This high-level
structure served as the foundation for our simulation experiments. Full
methodological details, refer to the \hyperref[sec-methods]{Methods}
section.}

@@

We present the results in two parts. First, we report the causal
structures inferred from the HELIUS data at the levels of aggregate
depression, individual symptoms, and aggregated precariousness. Second,
we embed these findings in a simplified dynamical model to simulate how
alternative feedback structures shape the system's response to
stress-reducing interventions.

@@

\caption{\label{fig-intervention}System response to external
intervention across variations in internal structure.
\emph{\textbf{Panel A}: Mean levels of depression (\(D\)) and
precariousness (\(P\)) at time steps 100, 300, and 500, across models
that differ in the coupling coefficient \(\alpha_{DP}\), which
represents the strength of depression's influence on precariousness in
the dynamical model. The x‑axis shows intervention strength, defined as
the proportional reduction of financial stress (\(S\)) relative to its
baseline (with higher values indicating more external financial
support). Each line corresponds to a different model defined by a
specific \(\alpha_{DP}\) value, color-coded from low (purple) to high
(yellow). Models with lower \(\alpha_{DP}\) exhibit faster early
reductions in \(D\), while higher \(\alpha_{DP}\) leads to more
immediate declines in \(P\). \textbf{Panel B}: Full time trajectories
for two models---one with no coupling (\(\alpha_{DP} = 0\)) and one with
strong coupling (\(\alpha_{DP} = 0.65\))---illustrating system response
and recovery following intervention onset (step 300) and withdrawal
(step 1500). Depression responds more gradually and persistently in the
high \(\alpha_{DP}\) model, whereas precariousness adjusts more
quickly.}}

@@

This study combined constraint-based causal discovery and dynamical
modeling to investigate how depression and socioeconomic precariousness
interact and how their internal structure modulates responsiveness to
intervention. Consistent with prior work linking financial strain to
psychological distress\textsuperscript{13,14}, our analysis identified
financial stress as a key upstream driver of depression in both
aggregated and disaggregated models. This supports longstanding
hypotheses that economic stressors can act as modifiable entry points
into broader cycles of psychosocial risk.

At the symptom level, our analysis aligns with established theories of
psychopathology but provides a more fine-grained map of potential causal
relationships, generating concrete, testable hypotheses. It also points
to possible limitations of prevailing models, as some observed
symptom--symptom links may be shaped by shared confounders rather than
direct causal influence. This suggests that, in some contexts, broader
mechanisms or clusters of symptoms may provide more effective units of
analysis and intervention than individual symptoms alone. Within this
framework, sleep disturbance emerged as a potentially initiating
symptom, with directed connections to both symptom-level and
precariousness variables---mirroring longitudinal findings that identify
sleep disruption as an early warning signal for depressive episodes and
social withdrawal\textsuperscript{15,16}. Depressed mood, another core
symptom, occupied a central position in the network. Not only was it
influenced by upstream symptoms like anhedonia and guilt, but it also
showed directional influence on social precariousness---suggesting that
internal psychological states may actively shape external vulnerability.
The CCI algorithm further identified a potential feedback loop between
depressed mood and overall precariousness, resonating with social drift
theories in depression, which posit that declining mental health can
lead to social exclusion, job instability, and housing insecurity,
reinforcing depressive cycles\textsuperscript{17,18}.

In contrast, social precariousness itself appeared more often as an
effect than a cause, receiving arrows from symptoms like depressed mood
and guilt but rarely sending them. This asymmetry supports prior
observations that indicators of social vulnerability often reflect
downstream consequences of poor mental health, though the relationships
are likely bidirectional over time\textsuperscript{13}. Although our
findings are based on Amsterdam-specific data, they align with research
in other high-income urban settings, where structural inequalities such
as poverty, unemployment, and housing insecurity consistently predict
poor psychological outcomes\textsuperscript{19,20}. While local contexts
may alter specific pathways, these parallels suggest that similar
mechanisms could operate elsewhere, though replication in other
populations is needed to assess robustness.

Taken together, these findings yield actionable insights for mental
health intervention. Symptoms like sleep disturbance, which demonstrate
early and broad influence across both symptom and precariousness
networks, may serve as effective targets for prevention---potentially
halting the progression into more entrenched depressive states or
worsening precariousness. Similarly, targeting symptoms such as
depressed mood, which appear embedded in potential feedback loops
linking psychological and social domains, may offer strategic leverage
for breaking mutually reinforcing cycles of mental decline and social
isolation. Importantly, interventions targeting such nodes may have
ripple effects across domains. For instance, evidence suggests that
treating core depressive symptoms can improve social functioning,
employment engagement, and housing stability---particularly when
combined with social or financial supports\textsuperscript{21,22}. These
dual benefits underscore the value of integrated, cross-domain
strategies that simultaneously address internal symptoms and external
stressors.

Our results also highlight why single-point interventions may fail to
produce lasting change. The broader causal structure reveals multiple
interlocking feedback loops spanning both psychological symptoms and
structural risk factors, which may re-amplify vulnerability even after
temporary relief. Supporting this, recent simulation work
by\textsuperscript{23} shows that feedback loops, even when modest in
strength, can substantially diminish the long-term impact of isolated
interventions unless multiple reinforcing links are simultaneously
disrupted. Our own simulations support this view: while all models
converged to similar long-term outcomes under sustained support,
differences in internal structure---specifically, the strength of the
link from precariousness to depression---produced systematic variation
in how quickly and through which pathways systems responded to
intervention. Stronger coupling slowed recovery and prolonged
depression, showing that internal architecture can be as important as
external input. This supports growing calls for systems-oriented mental
health perspectives\textsuperscript{24}. Rather than assuming uniform
responses, more durable outcomes may require (1) targeting early-acting
symptoms, (2) sustaining support where recovery is slow, and (3)
combining structural and psychological interventions. Tailoring
strategies to match the system's architecture can improve effectiveness,
particularly in populations with layered vulnerabilities. In this light,
understanding causal architecture is not merely theoretical---it is
foundational to designing more effective and equitable interventions.

While the FCI and CCI algorithms are designed to recover causal
structure from cross-sectional data---including feedback and latent
confounding---many edge directions in our results remain ambiguous,
particularly in links between individual symptoms and precariousness
factors. These inconsistencies likely reflect limited statistical signal
in high-dimensional settings, where large conditioning sets reduce power
and small effect sizes are harder to detect. Moreover, our findings
showed sensitivity to the choice of conditional independence (CI) test.
The Gaussian CI test, which assumes linearity and Gaussian noise, often
produced denser graphs, likely due to overestimating weak or noisy
linear associations. In contrast, RCoT, a nonparametric alternative,
avoids these assumptions but relies on a Gaussian radial basis function
(RBF) kernel to assess dependence. While flexible in capturing nonlinear
relationships, this kernel is optimized for smooth, continuous variables
and may underperform with discrete or ordinal data. Its similarity
estimates can be unreliable when applied to variables with
discontinuities, potentially leading to missed dependencies. This is
particularly relevant in the HELIUS dataset, where some variables are
ordinal or mixed-type\textsuperscript{25}. To address this, future work
should consider adapting kernel-based CI tests to better handle mixed
data structures---for example, by employing hybrid kernels that
explicitly combine continuous and discrete similarity measures. This
could improve the reliability of nonparametric causal discovery in
settings with heterogeneous variable types.

Although the CCI algorithm is explicitly designed to detect cycles,
aside from the connection between precarity and depression in the
aggregate precarity setting, we found little clear evidence of feedback:
most candidate loops appeared only as bidirectional edges, which may
reflect latent confounding or statistical imprecision. This could point
to either an absence of strong feedback at the symptom and the
precariousness factor levels or limitations in the sensitivity of
full-graph discovery approaches applied to densely interconnected
variables. More focused strategies, such as local structure tests or
theory-guided subgraph modeling, may help resolve these ambiguities.
Moreover, our analyses relied solely on observational data from the
Amsterdam population. Some structural uncertainties, particularly those
involving latent confounding or bidirectional influence, may ultimately
require interventional or semi-interventional designs. Extensions such
as LLC\textsuperscript{26}, NODAGS-Flow\textsuperscript{27}, or
Bicycle\textsuperscript{28}, which integrate observational and
interventional data, could be especially valuable where such data are
available.

Finally, the structure of our dynamical model introduces important
limitations. The model focused on a restricted set of variables
(depression, social precariousness, and financial stress). This design
was intentionally simplified to isolate core mechanisms of interest, but
it necessarily omits other relevant domains of precariousness,
additional stressors, and potential unintended consequences of
intervention. In addition, while the linear formulation offered
analytical tractability and transparent calibration, it likely
oversimplifies the true dynamics of depression and socioeconomic
precariousness. A growing body of evidence suggests that psychosocial
systems are fundamentally nonlinear. Empirical and theoretical work
points to phenomena such as threshold effects, where improvements only
occur after a tipping point is crossed\textsuperscript{29,30};
saturation, where the impact of support diminishes as basic needs are
met\textsuperscript{31}; and path dependence, where past adversity
shapes sensitivity to future stressors\textsuperscript{32}. These
behaviors are fundamentally incompatible with linear dynamics, yet they
may be essential for understanding why some individuals remain trapped
in high-risk states while others recover more easily. Future modeling
efforts should therefore incorporate mechanisms such as saturating
input-response functions (e.g., sigmoidal form) and state-dependent
feedback (where coupling strength varies by system state). Incorporating
such features would allow models to capture not just average effects but
also the critical conditions under which meaningful change becomes
possible---clarifying whether some subgroups require forceful,
threshold-crossing interventions, while others may benefit from
sustained, moderate support.

As public health research has long shown, social and psychological
systems do not always respond passively to external inputs; internal
constraints can limit the impact of even well-intentioned
policies\textsuperscript{33,34}. By using causal discovery and
computational modeling, researchers and policymakers can better identify
these structural constraints and tailor interventions accordingly. More
broadly, this study offers a generalizable framework for investigating
complex mental health dynamics. Causal discovery supports principled
hypothesis generation from observational data, while dynamical modeling
enables exploration of how hypothesized structures behave under external
change. Together, these tools move us beyond static descriptions toward
mechanistic insight with direct relevance for real-world mental health
policy and intervention design. Future research should build on this
foundation by developing richer, more nuanced dynamical models grounded
in longitudinal and interventional data. Incorporating time, multiple
domains of precarity, and adaptive feedback processes would help assess
whether the mechanisms proposed manifest in actual mental health
trajectories. Ultimately, we hope this study encourages further
integration of causal structure learning and systems modeling in efforts
to understand and improve mental health outcomes under conditions of
socioeconomic precariousness.

\section{Methods}\label{sec-methods}

We use data from the HELIUS (HEalthy LIfe in an Urban Setting)
study\textsuperscript{11,12}. It captures the diverse population of the
city of Amsterdam by including the six main ethnic groups and provides
comprehensive health and lifestyle data, including depressive symptoms
as measured by the PHQ-9\textsuperscript{35}. To operationalize
indicators of precariousness, we draw on the framework outlined in
previous research\textsuperscript{9} and select a set of relevant
variables. To ensure a robust representation of precariousness in our
causal discovery models, we conducted exploratory analyses to identify
consistent and meaningful data-driven factor structures. These analyses
led to the identification of five distinct precariousness factors: three
reflecting structural domains---employment, housing, and social
relationships---and two capturing recent life stressors---financial and
relational. Each factor is composed of multiple observed variables, as
outlined below. Additional details on the HELIUS study and the factor
extraction process are provided in the
\hyperref[sec-appendix]{Appendix}.

Following data cleaning and normalization procedures, the analytical
sample comprises 21,628 individuals. In addition to the five identified
precariousness factors, we computed an overall precariousness score by
summing these five factor scores, thereby capturing a composite measure
of accumulated precariousness across domains. Depression is represented
using the PHQ-9\textsuperscript{36}, both as a total sum score and as
individual symptom scores. In the subsequent causal discovery analysis,
we explore the relationship between depression and precariousness using
both aggregated scores and their disaggregated components. See
Figure~\ref{fig-dist} for the overall distributions of the variables
used in the analysis.

\includegraphics[width=0.9\textwidth,height=\textheight]{draft_v3_files/figure-pdf/fig-dist-1.pdf}

To uncover directional relationships between precariousness and
depression, we use constraint-based causal discovery algorithms that
allow for feedback loops and latent confounding. Specifically, we apply
the \emph{Fast Causal Inference} (FCI) and \emph{Cyclic Causal
Inference} (CCI) algorithms, both of which can detect cycles and
non-acyclic structures\textsuperscript{37,38}. For reference, we also
include the \emph{PC} algorithm, one of the most widely used methods for
acyclic causal discovery\textsuperscript{39}. While our full analysis
includes results from three causal discovery algorithms---FCI, CCI, and
PC---we focus on FCI in the main text, as it offers a strong balance
between flexibility and theoretical soundness. FCI accommodates latent
confounding and potential feedback, making it well-suited for the
complexity of our data\textsuperscript{37}. The PC algorithm is included
for reference but relies on more restrictive assumptions, such as
acyclicity and no unmeasured confounding. CCI, while capable of modeling
cycles explicitly, has its own theoretical limitations; for example, its
output may not fully preserve the d-separation properties of the true
causal graph\textsuperscript{38}. Results from both PC and CCI are
provided in the \hyperref[sec-appendix]{Appendix}
(Section~\ref{sec-cci}, Section~\ref{sec-pc}) for completeness and
comparison. For readers interested in detailed explanations of the
algorithms, edge interpretations, and graph types (e.g., PAG, MAAG,
CPDAG), as well as algorithmic assumptions and limitations, please refer
to the \hyperref[sec-appendix]{Appendix}
(Section~\ref{sec-causalprimer}).

A key challenge in applying these algorithms to the HELIUS dataset is
that many variables are non-Gaussian and likely exhibit nonlinear
relationships. While kernel-based conditional independence tests (such
as KCIT) are well-suited for capturing such complex dependencies, they
are computationally intensive---scaling quadratically with sample size
due to the need to invert large kernel matrices---making them
impractical for large datasets like HELIUS\textsuperscript{40}. To
address this, we use the Randomized Conditional Correlation Test (RCoT),
a nonparametric alternative that approximates kernel methods using
random Fourier features. This reduces computational complexity from
quadratic to linear in sample size, substantially lowering runtime while
preserving sensitivity to nonlinear
relationships\textsuperscript{41,42}. For further details, see
Section~\ref{sec-rcot}.

\section{Data Availibility}\label{data-availibility}

The data used in this study come from the HELIUS (HEalthy LIfe in an
Urban Setting) cohort. Due to privacy and ethical restrictions, the raw
HELIUS data are not publicly available. Researchers who wish to access
the data can submit a request to the HELIUS study team via the study
website (\url{https://heliusstudy.nl}), following the governance
procedures of the HELIUS Executive Board.

\section{Code Availibility}\label{code-availibility}

All analysis scripts used for causal discovery, model calibration, and
simulation are available at
\url{https://github.com/KyuriP/Precarity_HELIUS}. The repository
contains documented R code and instructions for reproducing the analyses
reported in this manuscript.
